
## Seeking Help

> help me help you

*Sometimes a second pair of eyeballs is all you need*

![Image credit: Intro to DS using R
Workshop](https://bcgov.github.io/ds-cop-intro-to-r/seeking-help-fr%20om-others-with-reprex.html)\](img/debugging.jpg){fig-align="center"
width="149"}

*Much of this information is also on the [FAQ](../faq.html) page*

> still accurate?

### Advice on asking for help

The key to receiving help from someone is for them to rapidly grasp your
problem. You should make it as easy as possible to pinpoint where the
issue might be.

Try to use the correct words to describe your problem. For instance, a
package is not the same thing as a library. Most people will understand
what you meant, but it can make things confusing for people trying to
help you. Be as precise as possible when describing your problem.

:warning: Don't let not knowing exactly how to describe your problem
prevent you from asking. Screenshots help tremendously!

**When asking someone for help try to** 1. Explain what thing you are
trying to do 2. Explain/show the code you wrote to try to do that thing
3. Explain/show your result, and if it's not obvious explain why you
feel it's not the correct result. (E.g. you expected the answer to be 5,
but instead it's 10. )

:warning: Don't spend more than 20 minutes banging your head on the wall
before you ask for help!

### Help from inside R Studio

#### Use the built-in RStudio help interface to search for more information on R functions

One of the fastest ways to get help is to use the RStudio help
interface. This panel by default can be found at the lower right hand
panel of RStudio. As seen in the screenshot, by typing the word `mean`,
RStudio tries to also give a number of suggestions that you might be
interested in. The description is then shown in the display window.

#### I know the name of the function I want to use, but I'm not sure how to use it

If you need help with a specific function, let's say `barplot()`, you
can type:

```{r, eval=FALSE, purl=FALSE}
?barplot
```

If you just need to remind yourself of the names of the arguments, you
can use:

```{r, eval=FALSE, purl=FALSE}
args(lm)
```

#### I want to use a function that does X, there must be a function for it but I don't know which one...

If you are looking for a function to do a particular task, you can use
the `help.search()` function, which is called by the double question
mark `??`. However, this only looks through the installed packages for
help pages with a match to your search request

```{r, eval=FALSE, purl=FALSE}
??kruskal
```

If you can't find what you are looking for, you can use the
[rdocumentation.org](http://www.rdocumentation.org) website that
searches through the help files across all packages available.

Finally, a generic Google or internet search "R \<task\>" will often
either send you to the appropriate package documentation or a helpful
forum where someone else has already asked your question.

#### I get an error message that I don't understand

Start by googling the error message. However, this doesn't always work
very well because often, package developers rely on the error catching
provided by R. You end up with general error messages that might not be
very helpful to diagnose a problem (e.g. "subscript out of bounds"). If
the message is very generic, you might also include the name of the
function or package you're using in your query.

If you check Stack Overflow, search using the `[r]` tag. Most questions
have already been answered, but the challenge is to use the right words
in the search to find the answers:
<http://stackoverflow.com/questions/tagged/r>

:warning: Development of R moves pretty fast. When at all possible, use
results from the past 1-2 years. Anything over 5 years old for packages
such as `ggplot`, `dplyr`, and `forcats` are likely obsolete.

### In Person help

-   Start by ask the person sitting next to you! Don't hesitate to talk
    to your neighbor, compare your answers, and ask for help.
-   Ask in the Math & Statistics [Discord
    server](https://discord.gg/ck497bJty5) - we have a dedicated channel
    just for this class but a lot of students in the server know how to
    use R.
-   Attend [Community
    Coding](https://www.csuchico.edu/datascience/community-coding.shtml).
    -   Drop in work session & dedicated space to work on coding
        projects.
    -   Collaborate with your peers and learn from experts.
    -   Some hours are held over Zoom

### Other Online

-   In RStudio go to `Help` --\> `Cheatsheets` or to this URL
    https://posit.co/resources/cheatsheets/
-   [Stack Overflow](http://stackoverflow.com/questions/tagged/r): if
    your question hasn't been answered before and is well crafted,
    chances are you will get an answer in less than 5 min. Remember to
    follow their guidelines on [how to ask a good
    question](http://stackoverflow.com/help/how-to-ask).
-   [Chico R Users Google
    Group](https://groups.google.com/forum/#!forum/chico-rug).
-   The [Posit Community](https://forum.posit.co/): it is read by a lot
    of people and is welcoming to new users.
-   If your question is about a specific package, see if there is a
    mailing list for it. Usually it's included in the DESCRIPTION file
    of the package that can be accessed using
    \`packageDescription("name-of-package:. You may also want to try to
    email the author of the package directly, or open an issue on the
    code repository (e.g., GitHub).

### Using AI tools

> write guidance here

### Written

If you're a book kinda person, there is plenty of help available as
well. Many have online versions or free PDF's.

-   R Markdown, the Definitive Guide:
    https://bookdown.org/yihui/rmarkdown/
-   R for Data Science https://r4ds.had.co.nz/
-   Cookbook for R http://www.cookbook-r.com/
-   R Graphics Cookbook (I use this all the time) -- Chapter 8 in the
    above link
-   The Art of R Programming https://nostarch.com/artofr.htm
-   R for... http://r4stats.com/
    -   Excel Users https://www.rforexcelusers.com/
    -   SAS and SPSS Users http://r4stats.com/books/r4sas-spss/
    -   STATA Users http://r4stats.com/books/r4stata/
