---
title: "Homework 2"
author: "Solutions"
date: today
format: 
  pdf: 
    tbl-pos: H
execute: 
  warning: false
  message: false
---

# Introduction

In this assignment you will practice summarizing variables, creating new variables, handling missing data, and managing factor variables.  You will work with two data sets from the `openintro` package

```{r}
library(forcats)
library(gtsummary)
ncbirths <- openintro::ncbirths
smoking <- openintro::smoking
```

# Part I: Working with Data Frames (NC Births)

## A. Summarizing variables

1. Calculate the `mean` and `median` of the father’s age (`fage`), removing missing values as needed.

```{r}
mean(ncbirths$fage, na.rm=TRUE)
median(ncbirths$fage, na.rm=TRUE)
```

2. Pregnancies typically last about 38 weeks. _Update_ the existing variable `weeks` so that any value greater than 38 is set to 38. That is, for all record where `weeks>38`, change the value of `weeks` to `<- 38`. Display the `summary` of `weeks` to confirm that the maximum value is now 38.


```{r}
ncbirths$weeks[ncbirths$weeks > 38] <- 38
summary(ncbirths$weeks)
```

3. Create a new logical variable called `missing_gained` that indicates whether the variable `gained` is missing. Show a table of your result. 

```{r}
ncbirths$missing_gained <- is.na(ncbirths$gained)
table(ncbirths$missing_gained)
```

4. Calculate the **proportion** of records with missing values in `gained` using **two different methods**.
```{r}
table(ncbirths$missing_gained) |> prop.table()
tbl_summary(ncbirths, include = "missing_gained")
```

5. Use the `ifelse` function to create a new variable called `term_status` where pregnancies with `weeks >= 37` are labeled `"term"` and pregnancies with `weeks < 37` are labeled `"preterm"`. Create a frequency table to verify your result.
```{r}
ncbirths$term_status <- ifelse(ncbirths$weeks>=37, "term", "preterm")
table(ncbirths$term_status)
```

# Part II: Working with Factors (Smoking Data)

1. Use `fct_count()` to examine the distribution of the variable `ethnicity`.
```{r}
fct_count(smoking$ethnicity)
```

2. Create a new factor variable `ethnicity_collapsed` by modifying the variable `ethnicity` such that:

- `"Refused"` and `"Unknown"` are dropped
- `"Asian"` and `"Chinese"` are combined into `"Asian"`
- all other levels remain unchanged  

Verify your recode using a two-way table comparing the old `ethnicity`and new variables.
```{r}
smoking$ethnicity_collapsed <- fct_collapse(smoking$ethnicity,
                                            Asian = c("Asian","Chinese"))

smoking$ethnicity_collapsed[smoking$ethnicity_collapsed %in% c("Refused", "Unknown")] <- NA
smoking$ethnicity_collapsed <- fct_drop(smoking$ethnicity_collapsed)
table(smoking$ethnicity, smoking$ethnicity_collapsed)
```

3. Using `ethnicity_collapsed`, create a new variable called `ethnicity_code` with the following labels: `"A"` for Asian, `"B"` for Black, `"M"` for Mixed, `"W"` for White. Display a frequency table using `tbl_summary` of the new variable.

```{r}
smoking$recode_ethnicity <- smoking$ethnicity_collapsed |>
        fct_recode("A"="Asian", "B"="Black","M"="Mixed", "W"="White")

tbl_summary(smoking, include = recode_ethnicity)
```

4. Using the frequencies from the table above, reorder the levels of `ethnicity_collapsed` from _least frequent to most frequent_.  Print a table of the reordered factor to confirm the new order.

```{r}
smoking$ethnicity_collapsed %>% fct_relevel("Mixed", "Black","Asian","White") %>% table()
```

5. Create a new factor variable called `nationality_lumped` from `nationality` that keeps the four most frequent nationalities and combines all remaining levels into a single category called "Other". Display a table of the new variable that includes both the frequency and percent (n%)

```{r}
smoking$nationality_lumped <- fct_lump_n(smoking$nationality, n = 4)
tbl_summary(smoking, include = nationality_lumped)
```


