---
title: "Working with Data Frames - Notes"
author: "Key"
date: today
format: pdf
execute: 
  message: false
  warning: false
---

:::{.callout-important title = "How to use the Notes files"}
This notes file follows along with the lesson materials. 
Your job is to fill out the missing code, often denoted with _____, 
and to complete the "you try it" type problems presented in the notes. 

Render this document regularly, ensuring that it is complete before you submit.
Before you submit, remove the `error: true` line in the YAML header then render. 
:::

# Setup

```{r}
library(ggplot2)               # replace the blank line with `ggplot2`
library(gtsummary)              
ncbirths <- openintro::ncbirths 
```


### Frequency Tables

You can create a basic frequency table by using the `table()` function. 

```{r}
table(ncbirths$lowbirthweight)
```

Relative frequencies (proportions or percentages) are calculated by putting the results of the `table` function inside the `prop_table` function. 

```{r}
prop.table(
  table(ncbirths$lowbirthweight)
)
```

### Summary Statistics

The function `summary()` prints out the five number summary, and includes the mean. This function also displays the number of missing values for that variable. 

```{r}
summary(ncbirths$visits)
```


### Fancy summary tables

The `gtsummary` package provides a single function `tbl_summary` to create a really nicely formatted summary table for both quantitative and categorical data types. 

```{r}
tbl_summary(ncbirths, 
            include = c(visits, lowbirthweight)
            )
```

The `statistic` argument can be used to change what values are displayed.
```{r}
tbl_summary(ncbirths, 
            include = c(visits, lowbirthweight), 
            statistic = list(
                all_continuous() ~ "{mean} ({sd})",
                all_categorical() ~ "{n} / {N} ({p}%)"
              )
            )
```

## Missing Data 

We can see 4 out of the first 6 values for the variable `fage` (fathers age) in the `ncbirths` data set are missing. 

```{r}
head(ncbirths$fage)
```


**Problem 1: `R` can't do arithmetic on missing data.** 

```{r}
mean(ncbirths$fage)
```

:::{.callout-tip title = "👉 Fix this error" icon=false}
Add the argument `na.rm=TRUE` to the `mean()` function inside the right hand parenthesis to calculate the mean after excluding missing values. 

```{r}
mean(ncbirths$fage, na.rm = TRUE)
```

:::


**Problem 2: Some plots will show `NA` as it's own category**

Sometimes this is fine, other times this is undesirable. We'll see later how we can adjust this plot to remove that column of NA. 

```{r}
ggplot(ncbirths, aes(premie)) + geom_bar()
```


### Identifying missing values 

To find out how many values in a particular variable are missing we can use several different approaches. 

#### Look at the raw data {.unnumbered} 

```{r}
head(ncbirths)
```

#### Look at data summaries {.unnumbered}  

Functions such as `table()` have a `useNA="always"` option to show how many records have missing values, and `summary()` will always show a column for NA. 

```{r}
table(ncbirths$habit, useNA="always")
summary(ncbirths$fage)
```

#### Use a logical statement {.unnumbered}

The function `is.na()` returns TRUE or FALSE for each element in the provided vector for whether or not that element is missing.  
```{r}
x <- c("green", NA, 3)
is.na(x)
```

```{r}
sum(is.na(ncbirths$fage))
```

There are 171 records in this data set where the age for the father is not present. 

:::{.callout-note title = "Negating `is.na()` to find the non-missing values"}
Recall we can use the `!` to negate a boolean argument. 
:::

```{r}
!is.na(x)
```


:::{.callout-tip title = "👉 Hide the NA bar" icon=false}
We can use this tactic to fix that barchart from above. Wrap `!is.na()` around the `premie` variable in the `ggplot` code to create a barchart that does not have a bar for missing values. 

```{r}
ggplot(ncbirths, aes(!is.na(premie))) + geom_bar()
```

:::



## Data management 

### Overwrite existing values

:::{.callout-tip title = "Example 1: Too low birthweight"}
Let's look at the numerical distribution of birthweight (in pounds) of the baby. 

```{r}
summary(ncbirths$weight)
```
:::

The value of 1 lb seems very low. The researchers you are working with decide that is a mistake and should be excluded from the data. We would then set all records where `weight=1` to missing. 

```{r}
ncbirths$weight[ncbirths$weight==1] <- NA
```

```{r}
min(ncbirths$weight, na.rm=TRUE)
```


:::{.callout-tip title = "👉 Your Turn" icon=false}
But what about other weights that aren't quite as low as 1, but still unusually low? 

* Write the code to set all birth weights less than 4 lbs (`<4`) to missing (NA).
* Then recalculate the mean to confirm your recode worked. 

```{r}
ncbirths$weight[ncbirths$weight < 4] <- NA
min(ncbirths$weight, na.rm=TRUE)
```
:::

### Creating new variables 

The following code creates a new variable `wtgain_mom` the weight gained by the mother, that is not due to the baby by subtracting `weight` from `gained`. 
```{r}
ncbirths$wtgain_mom <- ncbirths$gained - ncbirths$weight
```

To confirm this variable was created correctly, we look at the data contained in three variables in question. 

```{r}
head(ncbirths[,c('gained', 'weight', 'wtgain_mom')])
```

### Dichtomizing data 

Let's add a variable to identify if a mother in the North Carolina births data set was underage at the time of birth. Specifically Make a new variable `underage` on the `ncbirths` data set. If `mage` is under 18, then the value of this new variable is `underage`, else it is labeled as `adult`. 

```{r}
ncbirths$underage <- ifelse(ncbirths$mage < 18, "underage", "adult")
```

**Trust but Verify**


```{r}
table(ncbirths$underage, useNA="always")
```

Next let's check it against the value of `mage` itself. Let's look at all rows where mothers age is either 17 or 18 `mage %in% c(17,18)`, and only the columns of interest. 

```{r}
ncbirths[ncbirths$mage %in% c(17,18),c('mage', 'underage')]
```

## Chaining commands

:::{.callout-tip title = "👉Example: Frequency tables & summary statistics using the pipe" icon=false}

First stating the variable, then pipe in the summary function. 
```{r}
ncbirths$mature |> table() # instead of table(ncbirths$mature)
ncbirths$mage |> mean() #instead of mean(ncbirths$mage)
```
:::


